## The linreg function

This function `StatsChitran::linreg` involves fitting a [linear regression](https://en.wikipedia.org/wiki/Linear_regression) model to a multivariate case while returning a message in case of [multicollinearity](https://en.wikipedia.org/wiki/Multicollinearity)

The model used is a non-weighted [OLS](https://en.wikipedia.org/wiki/Ordinary_least_squares)

\newpage

### The univariate function  
Start with a simple fit for a $y = ax+b+\epsilon$ case where $\epsilon \sim N(0, \sigma^2)$. Obviously the model in this case would be $f(x) = ax+b$

Then let $(x_i, y_i) \,\,| \,\, i \in [1,n]$ be the dataset. Then the the objective cost function involving the [RSS](https://en.wikipedia.org/wiki/Residual_sum_of_squares) according to OLS becomes
$$
\begin{align}
\mathcal{L}(a, b) &= \sum_i \{y_i - ax_i -b\}^2\\
\end{align}
$$ 
For minimizing $\mathcal{L}(a, b)$, we need
$$\nabla \mathcal{L} = 
\begin{pmatrix}
	\partial \mathcal{L}/\partial a \\
	\partial \mathcal{L}/\partial b
\end{pmatrix} =
\begin{pmatrix}
0\\
0
\end{pmatrix}
$$

Hence,
$$
\begin{align}
	\frac{\partial \mathcal{L}}{\partial a} = 
	-2\sum_i\{y_i - ax_i -b\}x_i &= 0\\
	\implies a\sum_i x_i^2 + b\sum_i x_i &= \sum_i x_i y_i
\end{align}
$$

Additionally,
$$
\begin{align}
	\frac{\partial \mathcal{L}}{\partial b} = -2\sum_i\{y_i - ax_i -b\} &= 0\\
	\implies a\sum_i x_i + b\sum_i 1 &= \sum_i y_i
\end{align}
$$

Thus, in this case, estimating the coefficient pair $a,\, b$ is a simple case for solving for the matrix

$$
\begin{pmatrix}
	\sum_i x_i^2 &\sum_i x_i\\
	\sum_i x_i &\sum_i 1
\end{pmatrix}
\cdot
\begin{pmatrix}
	a\\
	b\\
\end{pmatrix}
=
\begin{pmatrix}
\sum_i x_i y_i\\
\sum_i y_i
\end{pmatrix}
$$

Avoiding multicollinearity simply becomes a restriction of imposed non singularity on the matrix  
$$
M =
\begin{pmatrix}
	\sum_i x_i^2 &\sum_i x_i\\
	\sum_i x_i &\sum_i 1
\end{pmatrix}
$$
In a strict theoretial sense that becomes 
$$\sum_i x_i^2 \ne \frac{1}{n}\bigg\{\sum_i x_i \bigg\}^2$$
That is only possible if all $x_i$ values are the same, hence the dataset would be reduced by a dimension of at least one with respect to the geometry that it is trying to fit itself to, basically the definition of a [singular matrix](https://en.wikipedia.org/wiki/Singular_matrix).

\newpage

### The multivariate function  
Now, if the issue is extended to the multivariate case, herein one would have $y = a_1 x_1 + a_2 x_2 + \cdots + a_m x_m + a_{m+1} + \epsilon$ with $\epsilon \sim N(0, \sigma^2)$. Obviously the linear fit in this case is $f(x)=a_1 x_1 + a_2 x_2 + \cdots + a_m x_m + a_{m+1}$. 

Thus the dataset in this case would be $(x_{1,i},\,\, x_{2,i}, \,\,\cdots x_{m,i},\,\, y_i )$ with $i \in [1,n]$

In this case the objective would be $\mathcal{L}(a_1, a_2, \cdots, a_{m+1})$  
$$
\begin{align}
\mathcal{L} &= \sum_i\{a_1 x_{1,i} + a_2 x_{2,i} + \cdots + a_m x_{m,i} + a_{m+1} - y_i\}^2 \\
&= \sum_{i=1}^n \bigg\{a_{m+1} - y_i + \sum_{k=1}^{m}a_k x_{k, i}  \bigg\}^2
\end{align}
$$

Thus for the gradient to be zero  
$$
	\nabla \mathcal{L} = 
	\begin{pmatrix}
	\partial\mathcal{L}/\partial a_1\\
	\partial\mathcal{L}/\partial a_2\\
	\vdots\\
	\partial\mathcal{L}/\partial a_m\\
	\partial\mathcal{L}/\partial a_{m+1}\\
	\end{pmatrix} = 
	\begin{pmatrix}
	0\\
	0\\
	\vdots \\
	0\\
	0
	\end{pmatrix}
$$

That implies that $\forall p \in [1, m]$  

$$
\begin{align}
\partial \mathcal{L}/\partial a_p = 2\sum_{i = 1}^n \{a_{m+1} - y_i + \sum_{k=1}^m a_k x_{k,i}\}x_{p,i} &= 0 \\
\implies a_{m+1}\sum_{i=1}^n x_{p,i} + \sum_{i=1}^n x_{p,i}\bigg\{\sum_{k=1}^m a_k x_{k,i} \bigg\} &= \sum_{i=1}^n x_{p,i}y_i\\ 
\implies \sum_{k=1}^m a_k \sum_{i=1}^n x_{k,i}x_{p,i} + a_{m+1}\sum_{i=1}^n x_{p,i} &= \sum_{i=1}^n x_{p,i}y_i
\end{align}
$$

This means that the first $m$ equations can be written down as shown below  
$$
\begin{align}
	a_1(\sum_{i=1}^n x_{1,i}x_{1,i})+a_2(\sum_{i=1}^n x_{2,i}x_{1,i}) + \cdots +a_m(\sum_{i=1}^n x_{m,i}x_{1,i}) + a_{m+1}\sum_{i=1}^n x_{1,i} &= \sum_{i=1}^n x_{1,i}y_i\\
	a_1(\sum_{i=1}^n x_{1,i}x_{2,i})+a_2(\sum_{i=1}^n x_{2,i}x_{2,i}) + \cdots +a_m(\sum_{i=1}^n x_{m,i}x_{2,i}) + a_{m+1}\sum_{i=1}^n x_{2,i} &= \sum_{i=1}^n x_{2,i}y_i\\
	a_1(\sum_{i=1}^n x_{1,i}x_{3,i})+a_2(\sum_{i=1}^n x_{2,i}x_{3,i}) + \cdots +a_m(\sum_{i=1}^n x_{m,i}x_{3,i}) + a_{m+1}\sum_{i=1}^n x_{3,i} &= \sum_{i=1}^n x_{3,i}y_i\\
	a_1(\sum_{i=1}^n x_{1,i}x_{4,i})+a_2(\sum_{i=1}^n x_{2,i}x_{4,i}) + \cdots +a_m(\sum_{i=1}^n x_{m,i}x_{4,i}) + a_{m+1}\sum_{i=1}^n x_{4,i} &= \sum_{i=1}^n x_{4,i}y_i\\
	a_1(\sum_{i=1}^n x_{1,i}x_{5,i})+a_2(\sum_{i=1}^n x_{2,i}x_{5,i}) + \cdots +a_m(\sum_{i=1}^n x_{m,i}x_{5,i}) + a_{m+1}\sum_{i=1}^n x_{5,i} &= \sum_{i=1}^n x_{5,i}y_i\\
	\vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad &= \qquad \vdots\\
	\vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad \vdots \qquad &= \qquad\vdots\\
	a_1(\sum_{i=1}^n x_{1,i}x_{m,i})+a_2(\sum_{i=1}^n x_{2,i}x_{m,i}) + \cdots +a_m(\sum_{i=1}^n x_{m,i}x_{m,i}) + a_{m+1}\sum_{i=1}^n x_{m,i} &= \sum_{i=1}^n x_{m,i}y_i\\
\end{align}
$$

However since in parameter space there are $m+1$ parameters $a_1, a_2, a_3, \cdots, a_m, a_{m+1}$ hence $m+1$ conditions are needed wherein the $(m+1)$th condition comes from setting the last minima $\partial \mathcal{L}/\partial a_{m+1} = 0$ as shown below  

$$
\begin{align}
	\partial \mathcal{L}/\partial a_{m+1} = 2\sum_{i=1}^n \bigg\{ a_{m+1} - y_i + \sum_{k=1}^m a_k x_{k,i}\bigg\} &= 0\\
	\implies \sum_{k=1}^m a_k\sum_{i=1}^n x_{k,i} + a_{m+1}\sum_{i=1}^n 1 &= \sum_{i=1}^n y_i
\end{align}
$$

In an expanded form this expands to the form shown below  
$$
a_1\sum_{i=1}^n x_{1,i} + a_2\sum_{i=1}^n x_{2,i} + \cdots + a_m\sum_{i=1}^n x_{m,i} + a_{m+1}\sum_{i=1}^n 1 = \sum_{i=1}^n y_i
$$

With $m+1$ linear conditions in the parameters from $a_1$ to $a_{m+1}$, the $(m+1) \times (m+1)$ matrix representation of $[M]$ can be shown below  

$$
[M] = 
\begin{pmatrix}
\sum_{i=1}^n x_{1,i}^2 &\sum_{i=1}^n x_{2,i}x_{1,i} &\cdots &\sum_{i=1}^n x_{m,i}x_{1,i} &\sum_{i=1}^n x_{1,i}\\
\sum_{i=1}^n x_{1,i}x_{2,i} &\sum_{i=1}^n x_{2,i}^2 &\cdots &\sum_{i=1}^n x_{m,i}x_{2,i} &\sum_{i=1}^n x_{2,i}\\
\sum_{i=1}^n x_{1,i}x_{3,i} &\sum_{i=1}^n x_{2,i}x_{3,i} &\cdots &\sum_{i=1}^n x_{m,i}x_{3,i} &\sum_{i=1}^n x_{3,i}\\
\sum_{i=1}^n x_{1,i}x_{4,i} &\sum_{i=1}^n x_{2,i}x_{4,i} &\cdots &\sum_{i=1}^n x_{m,i}x_{4,i} &\sum_{i=1}^n x_{4,i}\\
\sum_{i=1}^n x_{1,i}x_{5,i} &\sum_{i=1}^n x_{2,i}x_{5,i} &\cdots &\sum_{i=1}^n x_{m,i}x_{5,i} &\sum_{i=1}^n x_{5,i}\\
\vdots &\vdots &\ddots &\vdots &\vdots\\
\sum_{i=1}^n x_{1,i}x_{m,i} &\sum_{i=1}^n x_{2,i}x_{m,i} &\cdots &\sum_{i=1}^n x_{m,i}^2 &\sum_{i=1}^n x_{m,i}\\
\sum_{i=1}^n x_{1,i} &\sum_{i=1}^n x_{2, i} &\cdots &\sum_{i=1}^n x_{m,i} &\sum_{i=1}^n 1
\end{pmatrix}
$$

Similarly, the $(m+1) \times 1$ coefficient matrix $[C]$ can be drawn as shown below.

$$
[C] =
\begin{pmatrix}
a_1\\
a_2\\
a_3\\
\vdots\\
a_{m}\\
a_{m+1}
\end{pmatrix}
$$

To be able to complete the definition, the result vector also has to be defined as a $(m+1) \times 1$ matrix $[B]$ as shown below

$$
[B] = 
\begin{pmatrix}
\sum_{i=1}^n x_{1,i}y_i\\
\sum_{i=1}^n x_{2,i}y_i\\
\sum_{i=1}^n x_{3,i}y_i\\
\vdots\\
\sum_{i=1}^n x_{m,i}y_i\\
\sum_{i=1}^n y_i\\
\end{pmatrix}
$$

With these definitions structured, now the relationship can be established  
$$
[M]\cdot [C] = [B]
$$

Obviously the non-singularity of $[M]$ is necessary in order not to reduce its rank else the situation of multicollinearity can arise.

\newpage

### Matrix representation

Additionally, a better representation of the matrixes $[M]$ and $[B]$ can be created. For this purpose, each datacolumn which signify the independent variables $x_1, x_2, \cdots, x_m$ can be treated as vectors $\mathbf{X}_1, \mathbf{X}_2, \cdots, \mathbf{X}_m$.  
An additional vector $\mathbf{1}$ is defined in this regard.  
Each of these vectors are treated as possesing a length of $n$ since the data are $n$--observations long as defined previously.  

Thus the vectors $\mathbf{X}_r, \mathbf{Y}, \mathbf{1}$ are defined in the manner shown below.  

$$\mathbf{X_r} =
\begin{pmatrix}
x_{r,1}\\
x_{r,2}\\
x_{r,3}\\
\vdots\\
x_{r,n}
\end{pmatrix}
\quad
\mathbf{Y} = 
\begin{pmatrix}
y_1\\
y_2\\
y_3\\
\vdots\\
y_n
\end{pmatrix}
\quad
\mathbf{1} = 
\begin{pmatrix}
1\\
1\\
1\\
\vdots\\
1
\end{pmatrix}
$$ 

As a consequence of these definitions, 
$$
\begin{align}
\mathbf{X}_p \cdot \mathbf{X}_q &= \sum_{i=1}^n x_{p,i}x_{q,i}\\\
\mathbf{X}_p \cdot \mathbf{Y} &= \sum_{i=1}^n x_{p,i}y_i \\
\mathbf{X}_p \cdot \mathbf{1} &= \sum_{i=1}^n x_{p,i}\\
\mathbf{Y} \cdot \mathbf{1} &= \sum_{i=1}^n y_i\\
\mathbf{1} \cdot \mathbf{1} &= n\\
\end{align}
$$

Hence, the matrix definitions can be updated as shown  

$$[M] = 
\begin{pmatrix}
\mathbf{X}_1\cdot\mathbf{X}_1 &\mathbf{X}_1\cdot\mathbf{X}_2  &\cdots &\mathbf{X}_1\cdot\mathbf{X}_m &\mathbf{X}_1\cdot\mathbf{1}\\
\mathbf{X}_2\cdot\mathbf{X}_1 &\mathbf{X}_2\cdot\mathbf{X}_2  &\cdots &\mathbf{X}_2\cdot\mathbf{X}_m &\mathbf{X}_2\cdot\mathbf{1}\\
\mathbf{X}_3\cdot\mathbf{X}_1 &\mathbf{X}_3\cdot\mathbf{X}_2  &\cdots &\mathbf{X}_3\cdot\mathbf{X}_m &\mathbf{X}_3\cdot\mathbf{1}\\
\vdots &\vdots &\ddots &\vdots &\vdots\\
\mathbf{X}_m\cdot\mathbf{X}_1 &\mathbf{X}_m\cdot\mathbf{X}_2  &\cdots &\mathbf{X}_m\cdot\mathbf{X}_m &\mathbf{X}_m\cdot\mathbf{1}\\
\mathbf{1}\cdot\mathbf{X}_1 &\mathbf{1}\cdot\mathbf{X}_2
&\cdots &\mathbf{1}\cdot\mathbf{X}_m &\mathbf{1}\cdot\mathbf{1}
\end{pmatrix}
$$

$$[B] =
\begin{pmatrix}
\mathbf{X}_1\cdot\mathbf{1}\\
\mathbf{X}_2\cdot\mathbf{1}\\
\mathbf{X}_3\cdot\mathbf{1}\\
\vdots\\
\mathbf{X}_m\cdot\mathbf{1}\\
\mathbf{Y}\cdot\mathbf{1}\\
\end{pmatrix}
$$

$$[C] = [M]^{-1}[B]$$

\newpage

### Function Implementation

For the function's implementation in `R`, the function accepts a dataframe `df` as its argument.  
This occurs such that each individual column of `df` is each of the vectors $\mathbf{X}_p$ along with $\mathbf{Y}$ being the last column of the dataframe.

Hence,

`df[,1]` = $\mathbf{X}_1$  
`df[,2]` = $\mathbf{X}_2$  
`df[,3]` = $\mathbf{X}_3$  
`df[,m]` = $\mathbf{X}_m$  
`df[,m+1]` = $\mathbf{Y}$  

Thus, the argument `df` should be a $m+1$ column dataframe while the funtion returns vector $[C]^T$.

\newpage

### Examples


```{r linreg-example, echo=TRUE, message=TRUE, warning=TRUE, fig.width=6, fig.height=6, fig.pos="H"}
library(StatsChitran) #Call the relevant library

###Estimate and plot the univariate case
X <- seq(-10, 10, by=0.01)
Y <- 3*X + 7.47 + rnorm(length(X), sd = 5) #Add gaussian noise
df <- data.frame(X, Y) #Create the raw dataframe
coeff <- linreg(df) #Call the function
plot(X,Y, type = 'l', col=rgb(0,0,0,0.25)) #Plot the raw data first
lines(X, coeff[1]*X+coeff[2], col='red') #Plot the fit onto the data
coeff #Check the coefficients
```

```{=latex}
\FloatBarrier
\clearpage
\Needspace{8\baselineskip}
```



```{r linreg-example1, echo=TRUE, message=TRUE, warning=TRUE, fig.width=6, fig.height=6}
library(StatsChitran) #Call the relevant library
###Estimate the multivariate case
##Build the dataset##############
X1 <- seq(-10, 10, by=0.01)
X2 <- X1^2
X3 <- X2^1.5
X4 <- X3^2
Y <- 7.3*X1 + 4.7*X2 + 3*X3 + 1.3*X4 + 11 + rnorm(n=length(X1), sd = 50) #Add gaussian noise
df <- data.frame(X1, X2, X3, X4, Y)
##Build the dataset##############
coeff <- linreg(df) #call the function
coeff #check the coefficients
```

